---
title: "Mapping"
author: "Hanjo Odendaal"
date: "01 March 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = "999") 
# Ensure strings come in as character types
options(stringsAsFactors = FALSE)

# Install packages
# Note: you must have installed each of these packages before loading them
# Note 2: There may be some versioning issues with ggplot & ggmap. 
# Check out this stackoverflow thread http://bit.ly/2lXHRFJ 
library(ggplot2)
library(ggmap)
library(maptools)
library(ggthemes)
library(rgeos)
library(broom)
library(dplyr)
library(plyr)
library(grid)
library(gridExtra)
library(reshape2)
library(scales)
library(dplyr)
```

# Create a theme for mapping
 
```{r}
# Define one that we will use for plots
plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 18,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey80", size = 0.1),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey80", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=8),
    axis.text = element_text(size=8),
    axis.title.x = element_text(hjust=1),
    axis.title.y = element_text(hjust=1),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"))
}

# And another that we will use for maps
mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 18,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey80", size = 0.1),
    strip.text = element_text(size=12),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey80", color = "white"),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"))
}

# Define some palettes
palette_9_colors <- c("#0DA3A0","#2999A9","#458FB2","#6285BB","#7E7CC4","#9A72CD","#B768D6","#D35EDF","#F055E9")
palette_8_colors <- c("#0DA3A0","#2D97AA","#4D8CB4","#6E81BF","#8E76C9","#AF6BD4","#CF60DE","#F055E9")
palette_7_colors <- c("#2D97AA","#4D8CB4","#6E81BF","#8E76C9","#AF6BD4","#CF60DE","#F055E9")
palette_1_colors <- c("#0DA3A0")
```

# Download and unzip shape files for San Francisco

```{r}
# Define the URL of the zipped shapefile
# URL <- "https://github.com/simonkassel/Visualizing_SF_home_prices_R/raw/master/Data/SF_neighborhoods.zip"
# # Download the shapefile to your working directory and unzip it.
# download.file(URL, "Data/SF_neighborhoods.zip")
# unzip("Data/SF_neighborhoods.zip")
# Read it into R as a spatial polygons data frame & plot
neighb <- maptools::readShapePoly("Data/SF_neighborhoods.shp")
plot(neighb)
```

# Distribution of house prices San Francisco

```{r}
# URL <- "https://raw.githubusercontent.com/simonkassel/Visualizing_SF_home_prices_R/master/Data/SF_home_sales_demo_data.csv"
# download.file(URL, "Data/SF_home_sales_demo_data.csv")

sf <- read.csv("Data/SF_home_sales_demo_data.csv")

ggplot(sf, aes(SalePrice)) + 
  geom_histogram(fill = palette_1_colors) +
  xlab("Sale Price($)") + ylab("Count") +
  scale_fill_manual(values = palette_1_colors) +
  plotTheme() + 
  labs(x="Sale Price($)", y="Count", title="Distribution of San Francisco home prices",
       subtitle="Nominal prices (2009 - 2015)", 
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel") +
  scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma)
```

Removing outliers and plotting violin plots
```{r}
sf <- 
  sf %>% 
  filter(!(abs(SalePrice - mean(SalePrice)) > 2.5*sd(SalePrice)))
  

ggplot(sf, aes(x = SaleYr, y = SalePrice, fill = factor(SaleYr))) + 
  geom_violin(color = "grey50") +
  xlab("Sale Price($)") + ylab("Count") +
  scale_fill_manual(values = palette_7_colors) +
  stat_summary(fun.y=mean, geom="point", size=2, colour="white") +
  plotTheme() + theme(legend.position="none") +
  scale_y_continuous(labels = comma) +
  labs(x="Year",y="Sale Price($)",title="Distribution of San Francisco home prices",
       subtitle="Nominal prices (2009 - 2015); Sale price means visualized as points",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")
```

We’ll create a bounding box delineated by the neighborhood shapefile and then download the basemap

```{r}

# Define the bounding box
bbox <- neighb@bbox
 
# Manipulate these values slightly so that we get some padding on our basemap between the edge of the data and the edge of the map
sf_bbox <- c(left = bbox[1, 1] - .01, bottom = bbox[2, 1] - .005, 
             right = bbox[1, 2] + .01, top = bbox[2, 2] + .005)
# Download the basemap
basemap <- get_stamenmap(
  bbox = sf_bbox,
  zoom = 13,
  maptype = "toner-lite")
 
# Map it
bmMap <- ggmap(basemap) + mapTheme() + 
  labs(title="San Francisco basemap")
bmMap
```

```{r}
ggmap(basemap) + 
  geom_point(data = sf, aes(x = long, y = lat, color = SalePrice), 
             size = .25, alpha = 0.6) +
  facet_wrap(~SaleYr, scales = "fixed", ncol = 4) +
  coord_map() +
  mapTheme() + theme(legend.position = c(.85, .25)) +
  scale_color_gradientn("Sale Price", 
                        colors = palette_8_colors,
                        labels = scales::dollar_format(prefix = "$")) +
  labs(title="Distribution of San Francisco home prices",
       subtitle="Nominal prices (2009 - 2015)",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")
```

Filter on 2 years and increase size if points 

```{r}


  ggmap(basemap) + 
  geom_point(data = subset(sf, sf$SaleYr == 2015 | sf$SaleYr == 2009), 
             aes(x = long, y = lat, color = SalePrice), 
             size = 1, alpha = 0.75) +
  facet_wrap(~SaleYr, scales = "fixed", ncol = 1) +
  coord_map() +
  mapTheme() +
  scale_color_gradientn("Sale Price", 
                        colors = palette_8_colors,
                        labels = scales::dollar_format(prefix = "$")) +
  labs(title="Distribution of San Francisco home prices",
       subtitle="Nominal prices (2009 & 2015)",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")

```

The price appreciation is readily apparent. Let’s zoom into just one neighborhood in the Mission District, where economic and cultural change are reported almost daily. First we’ll create a new data frame of just sales in the ‘Inner Mission Neighborhood’ and readjust the basemap. Then we’ll build a facetted time series map of sales.

```{r}
#Subset sales for the "Inner Mission neighborhood"
missionSales <- 
  sf %>% 
  filter(Neighborhood == "Inner Mission")

#Create a new basemap at the appropriate scale
centroid_lon <- median(missionSales$long)
centroid_lat <- median(missionSales$lat)
missionBasemap <- get_map(location = c(lon = centroid_lon, lat = centroid_lat), 
                          source = "stamen", maptype = "toner-lite", zoom = 15)

#Create a facet map by year
  ggmap(missionBasemap) + 
  geom_point(data = missionSales, aes(x = long, y = lat, color = SalePrice), 
             size = 2) +
  facet_wrap(~ SaleYr, scales = "fixed", ncol = 4) +
  coord_map() +
  mapTheme() + theme(legend.position = c(.85, .25)) +
  scale_color_gradientn("Sale Price", 
                        colors = palette_8_colors,
                        labels = scales::dollar_format(prefix = "$")) +
  labs(title="Distribution of Mission District home prices",
       subtitle="Nominal prices (2009 - 2015)",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")

```


```{r}
# We'll start by summarizing our existing sales data frame
sf.summarized <- ddply(sf, c("Neighborhood", "SaleYr"), summarise, 
                       medianPrice = median(SalePrice),
                       saleCount = length(SaleYr),
                       sdPrice = sd(SalePrice),
                       minusSd = medianPrice - sdPrice,
                       plusSD = medianPrice + sdPrice,
                       .progress = "text")

sf.summarized <- 
  sf %>% ungroup %>% 
  dplyr::group_by(Neighborhood, SaleYr) %>% 
  summarise(
    medianPrice = median(SalePrice),
    saleCount = length(SaleYr),
    sdPrice = sd(SalePrice),
    minusSd = medianPrice - sdPrice,
    plusSD = medianPrice + sdPrice 
  )

 
# Now we'll calculate the average annual home sale count for each neighborhood
yearly_sales <- 
  sf.summarized %>% 
  dplyr::group_by(Neighborhood) %>% 
  summarise(avg.yearly.sales = mean(saleCount))
 
# Then we will use a left join to join the average sale count data frame to the original summarized data frame. 
sf.summarized <- left_join(sf.summarized, yearly_sales, by = "Neighborhood")
 
# Next, we'll calculate the % change in neighborhood median home value. In order to do this we will need to reshape the data 
# again using the dcast function in the package reshape2. 
medByYear <- dcast(sf.summarized, Neighborhood ~ SaleYr, value.var = "medianPrice")

sf.summarized %>% 
  ungroup %>% 
  select(Neighborhood, SaleYr, medianPrice) %>% 
  tidyr::spread(SaleYr, medianPrice)
# Check out the reshaped data frame
head(medByYear)
 
# We now have a data frame in which each row is a neighborhood, each column is a year and the values are the corresponding median prices. 
# Now we can easily calculate the % change from 2009 to 2015.
medByYear <- 
  medByYear %>% 
  mutate(pctChange = (`2015` - `2009`) / `2009`)


# And join it back to our sf.summarized dataset
sf.summarized <- left_join(sf.summarized, medByYear[,c("Neighborhood", "pctChange")], 
                           by = "Neighborhood")
 
# Some neighborhoods have very low annual sale counts. 
# We will remove these neighborhoods from the dataset by converting them to NA.
sf.summarized$pctChange <- ifelse(sf.summarized$avg.yearly.sales < 10, NA, 
                                  sf.summarized$pctChange)
 
# Remember the neighborhood shapefile we imported at the beginning? We must now convert it to a format that ggplot understands.
neighb.tidy <- tidy(neighb, region = c('nbrhood'))
# Look at the resulting data frame to see how it has been transformed
head(neighb.tidy)
 
# Create an identical field to 'id' but with a name that will allow us to join the data frame to our summarized price data
neighb.tidy$Neighborhood <- neighb.tidy$id
 
# Now we're going to join these data frames together so that when we map the neighborhood polygons we can symbolize them using the summary
# stats we created
sf.summarized_tidy <- dplyr::left_join(sf.summarized, neighb.tidy, by = "Neighborhood", match = "all")
```


```{r}
ggmap(basemap) +
  geom_polygon(data = sf.summarized_tidy, 
               aes(x = long, y = lat, group = group, fill = medianPrice), 
               colour = "white", alpha = 0.75, size = 0.25) + 
  scale_fill_gradientn("Neighborhood \nMedian \nSale Price", 
                       colors = palette_8_colors,
                       labels = scales::dollar_format(prefix = "$")) +
  mapTheme() + theme(legend.position = c(.85, .25)) + coord_map() +
  facet_wrap(~SaleYr, nrow = 2) +
  labs(title="Median home price by neighborhood, San Francisco ",
       subtitle="Nominal prices (2009 - 2015)",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")
```

```{r}
# Plot the percent change in neighborhood median home value over time
ggmap(basemap) +
  geom_polygon(data = sf.summarized_tidy[which(sf.summarized_tidy$SaleYr == 2015), ], 
               aes(x = long, y = lat, group = group, fill = pctChange), 
               colour = "white", alpha = 0.75, size = 0.25) + 
  coord_map() +
  scale_fill_gradientn("% Change", colors = palette_8_colors,
                       labels = scales::percent_format()) +
  mapTheme() + 
  theme(legend.position = "bottom", 
        legend.direction = "horizontal", 
        legend.key.width = unit(.5, "in")) +
  labs(title="Percent change in median home prices, San Francisco",
       subtitle="Nominal prices (2009 - 2015)",
       caption="Source: San Francisco Office of the Assessor-Recorder\nNeighborhoods with too few annual transactions are withheld\n@KenSteif & @SimonKassel")

```

A picture begins to emerge when we look at percent change. It’s clear that wealthy areas around The Marina, Russian Hill and Nob Hill did not change much while areas south changed dramatically. Let’s explore the rate of change by generating some time series plots for the highest appreciating neighborhoods.
First, a bit of data wrangling.

```{r}
# Lets look at the top 8 most appreciating neighborhoods.
topPctChange <- unique(sf.summarized$pctChange) %>% sort(decreasing = TRUE) %>% head(8)
 
# Well use these percentages to subset our neighborhoods data frame
sfForTimeSeries <- sf.summarized[which(sf.summarized$pctChange %in% topPctChange), ]
```


```{r}
time.series <- ggplot(sfForTimeSeries, aes(x = SaleYr, group=Neighborhood)) +
  geom_line(aes(y = medianPrice)) +
  geom_ribbon(aes(ymin = minusSd, ymax = plusSD, fill = Neighborhood), alpha = 0.75) +
  facet_wrap(~Neighborhood, scales = "fixed", nrow = 4) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  ylab("Neighborhood median home price") + xlab(NULL) +
  plotTheme() +
  theme(
    legend.position = "none",
    panel.spacing.y = unit(1, "lines")
  ) +
  scale_fill_manual(values=palette_8_colors) +
  labs(title="Time series for highest growth neighborhoods, San Francisco",
       subtitle="Nominal prices (2009-2015); Median; Ribbon indicates 1 standard deviation",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")
time.series
```

This plot shows how quickly home prices appreciated in the City’s fastest changing neighborhoods. It is always helpful to add a bit of geographical context to plots like this. Next, we’ll generate a map cutout and merge it with the time series plot.

```{r}
sampleNeighborhoods <- ggmap(basemap) + 
  geom_polygon(data = neighb.tidy, aes(x = long, y = lat, group = group), 
               colour = NA, fill="black", alpha = 1) +
  geom_polygon(data = neighb.tidy[which(neighb.tidy$Neighborhood %in% sfForTimeSeries$Neighborhood), ], 
               aes(x = long, y = lat, group = group, fill = Neighborhood), 
               colour = "black") +
  coord_map() +
  scale_fill_manual(values=palette_9_colors)+
  mapTheme() + 
  theme(legend.position = "bottom", 
        legend.direction = "horizontal", 
        legend.key.size = unit(0.15, "in"),
        legend.title=element_blank()) +
  guides(fill=guide_legend(ncol = 2))
sampleNeighborhoods
```

```{r}
# First well create a blank grob (or grid object) which will allow us to appropriately space out our plot below
blank <- grid.rect(gp = gpar(col="white"))

# Next, we will create an ordered list of the plots we want to use, including 'blank' 
gs <- list(time.series, blank, sampleNeighborhoods)

# Finally, we will denote a matrix that will dictate the layout of the plots. Each number
# in the matrix refers to a position in the list of plots.
lay <- rbind(c(1,1,2),
             c(1,1,2),
             c(1,1,3),
             c(1,1,3),
             c(1,1,2),
             c(1,1,2))

# Arrange the plot
grid.arrange(grobs = gs, layout_matrix = lay)
```

```{r}
# First we create a data frame of just 2009 and remove neighborhoods 
# that have an NA value for pctChange due to on insufficient sale volume
sf.2009 <- sf.summarized[which(sf.summarized$SaleYr == 2009), ] %>% na.omit()

# Then create the scatterplot using neighborhood name labels instead of points
change_scatterplot <- ggplot(sf.2009, aes(x = pctChange, y = medianPrice, label = Neighborhood)) + 
  geom_label(data = sf.2009[which(!sf.2009$pctChange %in% topPctChange),], 
             aes(label = Neighborhood), fill = "grey20", size = 2, color = "white") +
  geom_label(data = sf.2009[which(sf.2009$pctChange %in% topPctChange),], 
             aes(label = Neighborhood, fill = Neighborhood), size = 2, color = "white") +
  scale_fill_manual(values=pallete_9_colors) +
  geom_smooth(method = "lm", se = FALSE) +
  plotTheme() + theme(legend.position = "none") +
  scale_y_continuous(labels = dollar_format(prefix = "$")) + 
  scale_x_continuous(labels = percent, limits = c(min(sf.2009$pctChange - .04), max(sf.2009$pctChange + .025)) ) +
  labs(x="% Change", y="Median Sale Price (2009)",
       title="Change in home price as a function of initial price",
       subtitle="Median price; Change between 2009 - 2015",
       caption="Source: San Francisco Office of the Assessor-Recorder\n@KenSteif & @SimonKassel")
change_scatterplot
```

